---
title:  "Reproducible Research: Peer Assessment 1"
author: "Natália Faraj Murad"
date:   "9/23/2020"
output: 
  html_document:
    keep_md: true
---

```{r echo = FALSE, include = FALSE}
Sys.getlocale("LC_TIME")
Sys.setlocale(category="LC_TIME", locale="en_US.UTF-8")
```

## Loading and preprocessing the data
```{r}
library("data.table")
library(ggplot2)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'), method = "curl")
unzip("repdata%2Fdata%2Factivity.zip",exdir = "data")
```

## Reading csv Data into Data.Table. 
```{r}
activityData <- data.table::fread(input = "data/activity.csv")
```

## What is mean total number of steps taken per day?

```{r}
total_steps <- activityData[, c(lapply(.SD, sum, na.rm = FALSE)), .SDcols = c("steps"), by = .(date)] 
head(total_steps, 10)
```

Total number of steps by day histrogram

```{r}
ggplot(total_steps, aes(x = steps)) +
    geom_histogram(fill = "blue", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
total_steps[, .(mean_steps = mean(steps, na.rm = TRUE), median_steps = median(steps, na.rm = TRUE))]
```


## What is the average daily activity pattern?
Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r}
intervaldata <- activityData[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval)] 
ggplot(intervaldata, aes(x = interval , y = steps)) + geom_line(color="blue", size=1) + labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")
```

5-minute interval with the maximum number of steps?

```{r}
intervaldata[steps == max(steps), .(max_interval = interval)]
```



## Imputing missing values

Total number of missing values in the dataset

```{r}
activityData[is.na(steps), .N ]
```

Filling in all of the missing values in the dataset.

```{r}
# Filling in missing values with the median. 
activityData[is.na(steps), "steps"] <- activityData[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
```

New dataset with the missing data filled in.

```{r}
data.table::fwrite(x = activityData, file = "data/tidyData.csv", quote = FALSE)
```

Histogram of the total number of steps taken each day, mean and median total number of steps taken per day.

```{r}
# total number of steps taken per day
total_steps <- activityData[, c(lapply(.SD, sum)), .SDcols = c("steps"), by = .(date)] 
# mean and median total number of steps taken per day
total_steps[, .(mean_steps = mean(steps), median_steps = median(steps))]
ggplot(total_steps, aes(x = steps)) + geom_histogram(fill = "blue", binwidth = 1000) + labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

Type of Estimate | Mean_Steps | Median_Steps
--- | --- | ---
First Part (with na) | 10765 | 10765
Second Part (fillin in na with median) | 9354.23 | 10395

## Are there differences in activity patterns between weekdays and weekends?
New factor variable in the dataset with levels - “weekday” and “weekend” - indicating whether a given date is a weekday or weekend day.

```{r}
#New factor variable.
activityData <- data.table::fread(input = "data/activity.csv")
activityData[, date := as.POSIXct(date, format = "%Y-%m-%d")]
activityData[, `Day of Week`:= weekdays(x = date)]
activityData[grepl(pattern = "Monday|Tuesday|Wednesday|Thursday|Friday", x = `Day of Week`), "weekday or weekend"] <- "weekday"
activityData[grepl(pattern = "Saturday|Sunday", x = `Day of Week`), "weekday or weekend"] <- "weekend"
activityData[, `weekday or weekend` := as.factor(`weekday or weekend`)]
head(activityData, 10)
```

Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r}
activityData[is.na(steps), "steps"] <- activityData[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
intervaldata <- activityData[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval, `weekday or weekend`)] 
ggplot(intervaldata , aes(x = interval , y = steps, color=`weekday or weekend`)) + geom_line() + labs(title = "Avg. Daily Steps by Weektype", x = "Interval", y = "No. of Steps") + facet_wrap(~`weekday or weekend` , ncol = 1, nrow=2)
```
